---
title: "Modelling Species Distributions"
author: "Biology3579"
date: 2025-02-26
format: html
execute:
  warning: false
  message: false
editor: visual
---

## Set-up

**Important note:** *Comprehensive instructions, including all prerequisites and detailed guidelines for running this analysis and relevant code, are available in the README file of my [GitHub repository](https://github.com/Biology3579/SARSCoV2Assignment.git) for this project. Please refer to the README to ensure proper setup and usage.*

Before starting the  analysis, the environment needs to be properly configured to match the that which was used during the development of this project. This ensures that all dependencies and required libraries are installed correctly and that the analysis can be run reproducibly on different machines. 

The renv package is used to manage R package environments, ensuring that the correct versions of libraries are installed to avoid compatibility issues. The renv::restore() command reads the renv.lock file, which contains a snapshot of the package versions used during the initial setup. If any packages are missing or have mismatched versions, they will be automatically installed or updated to match the specified environment. This ensures that the project runs consistently across different systems.

```{r}
#| label: setup-renv
#| warning: false
#| message: false

# Restore the exact package versions used when the script was created
# This ensures that the same set-up is used as when the code was originally developed
renv::restore()
```

Once the environment is set up, the libraries that will be used throughout the analysis need to be loaded. The load_libraries() function is sourced from an external script (libraries.R) and ensures that all required libraries are loaded in one step. Please refer to this script for a more comprehensive description of each package used.
```{r}
#| label: load-libraries
#| warning: false
#| message: false

# Source the library loading function from the libraries file
# Detailed description of each library and its use can be found in the sourced file
source(here::here("functions", "libraries.R"))

# Load necessary libraries using the custom function
load_libraries()
```

```{r}
#| label: create-project-structure
#| include: false

# Further, a consistent directory structure is created to direct both raw and processed data, as well as output files to the correct place in the local directory
# This creates 'data/raw', 'data/processed', and 'output' directories if they don't already exist
sapply(c("data/raw", "data/processed", "output"), function(dir) {
    dir_path <- here(dir)
    if (!dir.exists(dir_path)) dir.create(dir_path, recursive = TRUE)
})

# This ensures the subfolder 'data/raw/climate' exists, creating it if necessary
climate_dir <- here("data", "raw", "climate")
if (!dir.exists(climate_dir)) {
  dir.create(climate_dir, recursive = TRUE)
}


```


## Introduction
As global climates change, ecosystems are facing increasingly unpredictable conditions that disrupt species interactions. For agricultural crops, this poses a significant challenge, as pests and pathogens spread more rapidly into new areas. In particular, climate change is expected to exacerbate the spread of invasive pests like the Red Palm Weevil (Rhynchophorus ferrugineus), which threatens the Date Palm (Phoenix dactylifera). Warmer temperatures, increased rainfall, and changing humidity levels could create more favorable conditions for the weevil, allowing it to expand into regions previously unaffected, thereby putting date palm cultivation at greater risk.

This growing problem highlights the need to understand how climate change will influence the interactions between crops and pests. In this analysis, Species Distribution Models (SDMs) are used to predict how the distributions of the Date Palm and the Red Palm Weevil might shift under future climate scenarios, helping to assess potential risks and inform management strategies for protecting date palms in the face of climate change. 

### Study Region
The analysis focuses on regions where the Date Palm (Phoenix dactylifera) is native, including North Africa, the Middle East, and parts of southern Europe. These areas are not only important for date palm cultivation but also for the species' natural occurrence. Given that the Red Palm Weevil (Rhynchophorus ferrugineus) is a major pest of date palms, it is likely to be found in these regions, as confirmed by preliminary analyses.

Additionally, a buffered extent around the defined study region was applied to ensure adequate coverage of surrounding habitats and to facilitate pseudo-absence sampling for model calibration.

```{r}
#| label: define-study-region
#| description: "Define the study extent and create a buffered region"

# Define study extent explicitly (longitude: -20 to 60, latitude: -5 to 45)
study_region <- terra::ext(-20, 60, -5, 45)

# Create a 5-degree buffered region for broader analysis
buffered_region <- extend(study_region, 5)
```

A global land shapefile (world_map) representing the global land areas, is used as a reference to create a study area mask. This mask helps focus the analysis on the specified geographic area, ensuring that only relevant land regions are included in the study.
```{r}
#| label: prepare-world-map
#| description: "Load global land shapefile and create a study area mask"

# Load global land shapefile for reference
world_map <- world(path = ".")

# Convert widened study region to raster for masking
study_area_mask <- terra::rasterize(world_map, rast(buffered_region))
```

## Species 1: Phoenix dactylifera (Date Palm)

### Data Preparation
Occurrence records for Phoenix dactylifera (and later for Rhynchophorus ferrugineus) were obtained from the Global Biodiversity Information Facility (GBIF), a large online database that compiles biodiversity data from various sources, including museums, research institutions, and citizen science initiatives []. GBIF provides information on species distributions, including location, date, and environmental context.

Before proceeding with the analysis, the raw dataset for Phoenix dactylifera occurrence records must first be processed to ensure it is appropriately structured. 
Accordingly, the raw data is cleaned and restructured, keeping only the most relevant variables and removing unnecessary columns. The data is then filtered to focus on the geographic region of interest and ocean points are excluded to ensure the analysis focuses on terrestrial occurrences of Phoenix dactylifera.

A full breakdown of these data processing steps is implemented in the loading_and_processing.R script, which is sourced in the chunks to ensure proper data handling and preparation for the analysis. 
```{r}
#| label: load-palm-data
#| description: "Load and process GBIF data for Phoenix dactylifera"

# Source data loading and processing functions
source(here("functions", "loading_and_processing.R"))

# Load raw GBIF data for date palm
palm_raw <- load_gbif_data("Phoenix dactylifera")
```

```{r}
#| label: process-palm-presence-data
#| message: false
#| description: "Clean and filter palm presence data"

# Source data loading and processing functions
source(here("functions", "loading_and_processing.R"))

# Create presence data for date palm
palm_presence_data <- palm_raw %>%
  clean_species_data() %>%    # Remove duplicates and missing values
  filter_coords() %>% # Filter coordinates to study region
  get_mainland_points # ...

# Save full distribution dataset
saveRDS(palm_presence_data, here("data", "processed", "palm_presence_data.rds"))

```

Since the presence data sourced from GBIF only includes locations where the species has been observed, pseudo-absence points are used to represent areas where the species is not present, creating a balanced dataset for species distribution modeling. These points help account for the environmental conditions where the species could potentially occur but has not been observed, allowing for more accurate modeling of its potential distribution. In this analysis, 500 pseudo-absence points are generated to complement the presence data and improve the model's predictive power.

```{r}
#| label: generate-palm-distribution-data
#| message: false
#| description: "Create distribution dataset with pseudo-absence points"

# Source data loading and processing functions
source(here("functions", "loading_and_processing.R"))

# Generate distribution data with pseudo-absence points
palm_distribution_data <- generate_pseudo_absence_data(palm_presence_data, num_points = 500)

# Save full distribution dataset
saveRDS(palm_distribution_data, here("data", "processed", "palm_distribution_data.rds"))
```

To visualise the distribution of Phoenix dactylifera, a map was created using observed occurrences and pseudo-absence points. The plotting functions for generating this map are sourced from the plotting.R script, where all the plotting functions are stored. For further details on the plotting function, please refer to this script.

```{r}
#| label: plot-palm-distribution
#| description: "Visualize date palm species distribution"

# Source plotting functions
source(here("functions", "plotting.R"))

# Create distribution plot
plot_species_distribution(palm_distribution_data, palm_colour, title = "Distribution of Date Palm")
```
_Figure 1: Map showing the distribution of Date Palm (Phoenix dactylifera) across the study region._

Figure 1 shows the geographic distribution of Date Palm within the study region. Purple points represent confirmed presence locations obtained from GBIF occurrence data, while grey points indicate pseudo-absence points. The distribution demonstrates the species' concentration across North Africa, the Middle East, and parts of southern Europe, highlighting its native range and primary ecological zones.

### Bioclimatic Variable Selection
The bioclimatic data is sourced from the WorldClim database, which provides high-resolution global climate data. The dataset includes various bioclimatic variables such as temperature, precipitation, and seasonality, which are crucial for understanding species distributions []. In this analysis, the specific bioclimatic data for the region of interest is either loaded from an existing file or downloaded and cropped to fit the study area.

```{r}
#| label: prepare-bioclim-data

# Define file path for bioclimatic data
bioclim_file_path <- here("data", "raw", "climate", "wc10.tif")

# Load existing raster or download and crop bioclimatic data
if (file.exists(bioclim_file_path)) {
  bioclim_data <- rast(bioclim_file_path)  # Load existing raster
} else {
  # Download, crop, and save bioclimatic data if not available
  bioclim_data <- crop(worldclim_global(var = "bio", res = 10, path = here("data", "raw", "climate")), buffered_region)
  writeRaster(bioclim_data, bioclim_file_path, overwrite = TRUE)
}
```

The selection of bioclimatic variables for Phoenix dactylifera (Date Palm) was based on previous literature showing that temperature extremes and moisture availability are the most critical factors affecting its distribution []. 
Specifically, the following variables were chosen:

- BIO5 (Max Temperature of the Warmest Month): Date Palms are sensitive to extreme heat, and this variable is crucial for assessing heat stress during the hottest periods, as noted in previous research.

- BIO6 (Min Temperature of the Coldest Month): This variable reflects the species' low tolerance for cold temperatures, ensuring suitability for warmer regions.

- BIO14 (Precipitation of the Driest Month): Selected to capture the species' tolerance to drought conditions, which is a key factor for Date Palm survival in arid environments.

- BIO16 (Precipitation of the Wettest Quarter): Chosen due to the species' sensitivity to wet stress, particularly during fruit ripening, which has been emphasized in earlier studies.Variables like BIO5 (Max Temperature of the Warmest Month), BIO6 (Min Temperature of the Coldest Month), BIO14 (Precipitation of the Driest Month), and BIO16 (Precipitation of Wettest Quarter) were chosen because they directly influence the species' survival, with temperature and moisture extremes being key limiting factors.

Although also identified as potentially relevant for predicting Phoenix dactylifera distributions, BIO10 (Mean Temperature of the Warmest Quarter) and BIO12 (Annual Precipitation) were excluded due to high collinearity with BIO5 and BIO16, and their lower relevance for capturing the critical temperature and moisture extremes affecting Date Palm's growth.

These decisions were guided by preliminary analyses conducted in the choosing_bioclims.R script (under supplementary scripts folder),  which assesses the correlations between selected bioclimatic variables. Please refer to this script for further detail.


```{r}
#| label: select-palm-bioclims
#| description: "Extract and prepare bioclimatic variables for palm species"

# Source processing functions
source(here("functions", "loading_and_processing.R"))

# Select and extract relevant bioclimatic variables
palm_model_data <- selected_bioclims(palm_distribution_data, 
  selected_bioclims = c("bio5", "bio6", "bio14", "bio16")
)

# Save processed dataset
saveRDS(palm_model_data, here("data", "processed", "palm_model_data.rds"))

```

### Model Fitting and Evaluation
The final model for Phoenix dactylifera was selected using AIC and AUC scores, as evaluated through model fitting and cross-validation (following ... recommendation). The process, detailed in the model_testing_script.R, involved fitting multiple Generalized Linear Models (GLMs) with selected bioclimatic variables.

- AIC-based Model Selection: Models were compared using AIC, and the model with the lowest AIC was selected for its balance between goodness of fit and simplicity.

- K-fold Cross-Validation (AUC): Models were further assessed using K-fold cross-validation, with AUC scores indicating the model with the best predictive performance.

The final model was chosen based on superior AIC and AUC scores, ensuring both statistical validity and predictive accuracy. For further details on the model selection, please refer to the model_testing_script.R.

```{r}
#| label: fit-palm-glm
#| description: "Fit generalized linear model for palm species distribution"

# Fit final GLM model with selected bioclimatic variables
final_palm_model <- glm(presence ~ bio5 + bio6 + bio14 + bio16,
    family = binomial(link = "logit"), data = palm_model_data)

# Print model summary
summary(final_palm_model)
```
_Palm Model Summary: GLM for date palm distribution _

The date palm distribution model reveals that occurrence probability decreases with increasing maximum temperatures and reduced precipitation, while slightly increasing with warmer winter temperatures. All bioclimatic variables are statistically significant, demonstrating the species' sensitivity to temperature and moisture extremes across its habitat range.

This GLM model was used to predict the climatic suitability of Date Palm across the study region. The output is plotted on Figure 2 and Figure 3.

```{r}
#| label: plot-palm-glm-probability-no-points
#| description: "Visualize GLM probability of palm occurrence without points"

# Source plotting functions
source(here("functions", "plotting.R"))

# Create probability plot without points
plot_glm_probability_raster(
  final_glm_model = final_palm_model,
  climate_data = bioclim_data,
  buffered_region = buffered_region,
  title = "GLM Probability of Palm Occurrence (Present)"
)
```
_Figure 2: Heat map showing the climatic suitability for date palm using the final_palm_model._

This heat map depicts the predicted climatic suitability for Phoenix dactylifera across North Africa, the Middle East, and southern Europe, using a color gradient representing occurrence probability from 0 to 1. Deep purple regions (indicating 75-100% probability) are concentrated in specific areas of North Africa, particularly along the western and northern coastal zones, and in concentrated pockets of the Arabian Peninsula. Yellows and greens (25-50% probability) dominate much of the broader region, including interior North African territories and parts of the Middle East. The plot highlights spatial variations in habitat suitability, with the most favourable conditions appearing in coastal and sub-coastal zones characterised by specific temperature and precipitation regimes. Areas of lowest probability (blue to teal) are evident in the northern Mediterranean regions and higher elevation areas, reflecting the species' sensitivity to temperature and moisture constraints.


```{r}
#| label: plot-palm-glm-probability-with-points
#| description: "Visualize GLM probability of palm occurrence with presence points"

# Create probability plot with presence points
plot_probability_raster_and_presence(
  palm_presence_data, 
  final_glm_model = final_palm_model, 
  buffered_region = buffered_region,
  species_colour = palm_colour, 
  title = "GLM Predicted Probability of Palm Occurrence"
)
```
_Figure 3: Heat map showing the date palm occurrence probability with presence points._

Figure 3 combines the GLM-predicted climatic suitability with actual presence points for date palm. Notably, the purple presence points strategically cluster in areas of higher predicted probability, demonstrating a strong overall correlation between the model's predictions and actual species distribution. However, the relationship is not perfectly uniform. While most points are concentrated in coastal regions and areas with favourable climatic conditions (yellow to green zones of 50-75% probability), there are notable discrepancies. Some presence points appear in areas of lower predicted suitability, particularly in the northern Mediterranean and transition zones, suggesting that local microclimatic factors or other biotic variables not captured by the model may influence species distribution. Conversely, some high-probability regions (dark blue and purple zones) show fewer presence points, indicating potential habitat suitability that has not yet been fully explored or documented. These variations highlight the limitations of relying solely on broad climatic variables to predict exact species occurrence.


## Species 2: Rhynchophorus ferrugineus (Red Palm Weevil)

### Data Preparation

The occurrence records for Rhynchophorus ferrugineus (Red Palm Weevil) were sourced from the Global Biodiversity Information Facility (GBIF), similarly to Phoenix dactylifera. 

The same data processing steps were applied to the raw weevil data - including cleaning the data, filtering for the study region, and generating pseudo-absence points - reusing the functions from the loading_and_processing.R script.

```{r}
#| label: load-weevil-data
#| description: "Load and process GBIF data for Rhynchophorus ferrugineus"

# Source data loading and processing functions
source(here("functions", "loading_and_processing.R"))

# Load raw GBIF data for red palm weevil
weevil_raw <- load_gbif_data("Rhynchophorus ferrugineus")
```

```{r}
#| label: process-weevil-presence-data
#| message: false

# Source data loading and processing functions
source(here("functions", "loading_and_processing.R"))

# Create presence data for red palm weevil
weevil_presence_data <- weevil_raw %>%
  clean_species_data() %>%       # Remove duplicates and missing values
  filter_coords() %>%            # Filter coordinates to study region
  get_mainland_points()          # Remove ocean points

# Save processed presence data
saveRDS(weevil_presence_data, here("data", "processed", "weevil_presence_data.rds"))
```

```{r}
#| label: generate-weevil-distribution-data
#| message: false

# Source data loading and processing functions
source(here("functions", "loading_and_processing.R"))

# Generate distribution data with pseudo-absence points
weevil_distribution_data <- generate_pseudo_absence_data(weevil_presence_data, num_points = 500)

# Save full distribution dataset
saveRDS(weevil_distribution_data, here("data", "processed", "weevil_distribution_data.rds"))
```

Figure 4 was created using observed occurrences and pseudo-absence points to visualise the distribution of Rhynchophorus ferrugineus. 
```{r}
#| label: plot-weevil-distribution
#| description: "Visualize red palm weevil species distribution"

# Source plotting functions
source(here("functions", "plotting.R"))

# Create distribution plot
plot_species_distribution(weevil_distribution_data, weevil_colour, title = "Distribution of Red Palm Weevil Presence & Pseudo-Absence")

```
_Figure 4: Map showing the distribution of Red Palm Weevil (Rhynchophorus ferrugineus) across the study region._

This map illustrates the spatial distribution of Red Palm Weevil occurrence across North Africa, the Middle East, and southern Europe. Red points represent confirmed presence locations, while grey points indicate pseudo-absence sampling sites. The Red Palm Weevil presence is notably concentrated along the Mediterranean coastal regions, with dense clusters in southern Spain, coastal North Africa (particularly Tunisia and Algeria), and the Levantine coast. The grey pseudo-absence points provide a comprehensive background, highlighting the potential habitat range beyond confirmed occurrence locations. The distribution suggests the weevil's preference for coastal and sub-coastal environments with specific climatic and ecological conditions conducive to its survival and spread.

### Bioclimatic Variable Selection

To select the appropriate bioclimatic variables for Rhynchophorus ferrugineus a brief reviw of prior research was done, which highlighted the importance of temperature extremes and moisture availability as key environmental factors influencing its habitat suitability [].

The following variables were selected for modeling the species' distribution:

-BIO2 (Mean Diurnal Temperature Range): This variable was chosen to reflect the weevil’s tolerance for temperature fluctuations, which is vital for its survival in warm climates, where high heat is prevalent.

-BIO6 (Min Temperature of the Coldest Month): This variable is critical as the Red Palm Weevil is highly sensitive to cold temperatures, which restricts its ability to expand into cooler regions. It plays a significant role in understanding the species' capacity to overwinter in colder environments.

-BIO16 (Precipitation of the Wettest Quarter): The Red Palm Weevil's sensitivity to excessive moisture during the wettest part of the year makes this variable important, particularly as wet conditions can negatively impact its survival during the reproductive phase.

-BIO17 (Precipitation of the Driest Quarter): Given the species’ preference for warm, humid environments, this variable captures the effect of dry periods that may limit its expansion into arid regions.

BIO1 (Annual Mean Temperature) was excluded from the analysis due to its high correlation with BIO2 (Mean Diurnal Temperature Range), since the annual mean temperature does not effectively capture the extreme conditions that are more relevant for the species’ survival. Instead, BIO6 was retained as it is more indicative of cold stress.

For further details on the variable selection process, please refer to the choosing_bioclims.R script

```{r}
#| label: select-weevil-bioclims
#| description: "Extract and prepare bioclimatic variables for weevil species"

# Process data for selected bioclimatic variables
weevil_model_data <- selected_bioclims(weevil_distribution_data, 
  selected_bioclims = c("bio2", "bio6", "bio16", "bio17")
)

# Save processed dataset
saveRDS(weevil_model_data, here("data", "processed", "weevil_model_data.rds"))
```

### Model Fitting and Evaluation
The final model was chosen based on superior AIC and AUC scores, ensuring both statistical validity and predictive accuracy. For further details on the model selection, please refer to the model_testing_script.R.

```{r}
#| label: fit-weevil-glm
#| description: "Fit generalized linear model for weevil species distribution"

# Fit final GLM model with selected bioclimatic variables
final_weevil_model <- glm(presence ~ bio2 + bio16 + bio17,
    family = binomial(link = "logit"), data = weevil_model_data)

# Print model summary
summary(final_weevil_model)
```
_Weevil Model Summary: GLM for red palm weevil distribution _

The Red Palm Weevil distribution model reveals two highly significant predictors with strong negative effects: temperature seasonality (bio2) and precipitation of the wettest quarter (bio16). The dramatic negative coefficient for temperature seasonality suggests the weevil is extremely sensitive to temperature fluctuations, while wet quarter precipitation also negatively impacts occurrence probability. Interestingly, precipitation of the driest quarter (bio17) shows no significant effect. 

This GLM model was used to predict the climatic suitability of Date Palm across the study region. The output is plotted on Figure 5 and Figure 6.

### Model Prediction and Visualization
The Red Palm Weevil's climatic suitability was predicted using the final GLM model and visualised on a map, showing high suitability in areas of the Middle East and North Africa.
```{r}
#| label: plot-weevil-glm-probability-no-points
#| description: "Visualize GLM probability of weevil occurrence without points"

# Source plotting functions
source(here("functions", "plotting.R"))

# Create probability plot without points
plot_glm_probability_raster(
  final_glm_model = final_weevil_model,
  climate_data = bioclim_data,
  buffered_region = buffered_region,
  title = "GLM Probability of Weevil Occurrence (Present)"
)
```
_Figure 5: Heat map showing the climatic suitability for red palm weevil using the final_palm_model._

Figure 5 illustrates the predicted climatic suitability for Rhynchophorus ferrugineus across North Africa, the Middle East, and southern Europe. It reveals a more restricted potential distribution compared to the Date Palm, with highest suitability strongly linked to coastal and Mediterranean climate regions. Areas of moderate to high probability are most evident in southern Spain, coastal North Africa, the Levantine coast, and parts of the Arabian Peninsula, reflecting the weevil's sensitivity to specific temperature and precipitation regimes. 

```{r}
#| label: plot-weevil-glm-probability-with-points
#| description: "Visualize GLM probability of weevil occurrence with presence points"

# Create probability plot with presence points
plot_probability_raster_and_presence(
  weevil_presence_data, 
  final_weevil_model, 
  buffered_region, 
  weevil_colour, 
  title = "GLM Predicted Probability of Weevil Occurrence"
)

```
_Figure 6: Heat map showing the red palm weevil occurrence probability with presence points._

This heat map integrates the GLM-predicted climatic suitability with actual presence points for the Red Palm Weevil. The red presence points seem to cluster in areas of higher predicted probability, primarily along the Mediterranean coastal regions. While most points align with yellow and green zones of moderate to high suitability (50-75% probability), there are notable discrepancies. Some presence points appear in areas of lower predicted probability, particularly in the northern Mediterranean transition zones, suggesting that local abiotic conditions or other unaccounted  variables influence species distribution. Conversely, some high-probability regions show fewer presence points, indicating potential habitat suitability not yet fully documented.

##  Question 3: Species Overlap Analysis

### Data Preparation

To determine the potential geographic regions where the two species may interact, a 10 km buffer was created around all confirmed presence records of Phoenix dactylifera. The buffer represents a potential interaction zone, where both species could be present. The overlap between the Red Palm Weevil and Date Palm occurrence zones was calculated by identifying the spatial intersection between the buffered Date Palm zone and the Red Palm Weevil presence data.

To process the spatial data, the presence records of both species were converted into simple feature (sf) objects, which allow for easy manipulation of geographic coordinates. The coordinates for both species' presence data were projected into a metric coordinate reference system (CRS) (EPSG: 3857) to allow for accurate distance and area calculations.

```{r}
#| label: process-species-overlap-data
#| warning: false
#| description: "Prepare spatial data for species overlap calculation"

# Convert presence data to sf objects
palm_sf <- st_as_sf(palm_presence_data, coords = c("lon", "lat"), crs = 4326)
weevil_sf <- st_as_sf(weevil_presence_data, coords = c("lon", "lat"), crs = 4326)

# Project to metric CRS 
crs_proj <- 3857
palm_proj <- st_transform(palm_sf, crs = crs_proj)
weevil_proj <- st_transform(weevil_sf, crs = crs_proj)

# Create buffer around palm presence points
palm_buffer <- st_buffer(palm_proj, dist = 10000)

# Calculate spatial intersection (weevils inside buffer)
weevil_overlap <- st_intersection(weevil_proj, palm_buffer)

```

To illustrate the spatial overlap between Date Palm and Red Palm Weevil distributions, a plot was created showing the presence points of both species, along with the overlapping areas. The plot highlights where the Red Palm Weevil occurrence points intersect with the Date Palm buffer zone, providing a representation of their potential co-occurrence.
```{r}
#| label: plot-species-overlap
#| description: "Visualize spatial overlap between palm and weevil distributions"

# Source plotting functions
source(here("functions", "plotting.R"))

plot_species_overlap_points(
  palm_points = palm_proj,
  weevil_points = weevil_proj,
  overlap_points = weevil_overlap,
  title = "Palm and Weevil Presence Overlap"
)

```
_Figure 5: Spatial co-occurrence of Date Palm and Red Palm Weevil._

This overlap map illustrates the spatial overlap between the distribution of date palm and red palm weevil across North Africa, the Middle East, and southern Europe. Purple points represent confirmed Date Palm presence locations, while red points correspond to Red Palm Weevil presence sites. Green points indicate the overlap between the two species, showing areas where both species occur within close proximity. As shown, the majority of overlap occurs in coastal and sub-coastal regions, particularly in parts of North Africa and the Middle East, where both species are likely to thrive. These areas are characterised by the appropriate climatic conditions for both species. The distribution of overlapping points suggests potential areas of increased risk for date palm cultivation, as these regions are more susceptible to red palm weevil infestations.

...The map also indicates areas where the weevil species occurs outside the palm buffer zone (red points outside the green area), which could signify the weevil's spread into regions where palms are not yet established. The overlap areas highlight potential ecological interactions between the two species, which could have implications for pest management strategies in these regions.

```{r}
#| label: calculate-overlap-percentage
#| description: "Calculate the percentage of weevil points within the palm buffer"

overlap_percentage <- mean(
  rowSums(st_within(weevil_proj, palm_buffer, sparse = FALSE)) > 0
) * 100

print(overlap_percentage)

```
The calculated overlap percentage of 28.8% indicates that approximately 28.8% of the red palm weevil presence records fall within the 10 km buffer around the Date Palm occurrence points. This relatively low overlap suggests that while there is some spatial co-occurrence between the two species, a substantial portion of the Red Palm Weevil's distribution lies outside the immediate range of Date Palms. This low overlap may reflect several factors, such as sampling biases, spatial barriers, or ecological constraints that prevent the two species from being more widely distributed together in the study region. It may also indicate that the Red Palm Weevil is capable of spreading into areas not yet inhabited by Date Palms, possibly due to dispersal dynamics or habitat suitability in regions outside the palms' natural range. The overlap percentage is useful for understanding the potential areas at risk for pest infestations in Date Palm regions. However, given the low overlap, further monitoring and analysis are needed to assess how the weevil might invade or affect Date Palm populations over time, particularly in regions where the two species do not yet occur together.

## Question 4

In this section, we examine whether the distribution of Phoenix dactylifera (Date Palm) is influenced by the presence of Rhynchophorus ferrugineus (Red Palm Weevil), while accounting for climatic variables. Specifically, we build a Generalised Linear Model (GLM) with palm presence as the response variable, considering both biotic factors (weevil presence) and environmental predictors (bioclimatic variables).

First, a 10 km buffer around each date palm occurrence point to investigate whether red palm weevil presence occurs near date palms. 

```{r}
# Convert Date Palm presence data to spatial format
palm_sf <- st_as_sf(palm_model_data, coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = st_crs(weevil_overlap))

# Create a 10 km buffer around Date Palm presence points
palm_buffer <- st_buffer(palm_sf, dist = 10000)  # 10 km buffer

```

Then the checked if the Red Palm Weevil presence occurred within the buffer zone surrounding the Date Palm presence points. We added a new variable, weevil_overlap, to the Date Palm dataset, which indicates whether each palm point is within the weevil presence zone.

```{r}
# Convert Red Palm Weevil presence data to spatial format
weevil_sf <- st_as_sf(weevil_presence_data, coords = c("lon", "lat"), crs = 4326)

# Transform to the same CRS as palm data for spatial operations
weevil_proj <- st_transform(weevil_sf, crs = 3857)

# Check if Red Palm Weevil occurs within the palm buffer
palm_model_data$weevil_overlap <- as.integer(
  lengths(st_intersects(palm_sf, palm_buffer)) > 0
)

# Create a binary overlap indicator
palm_model_data$weevil_overlap <- as.integer(lengths(st_intersects(palm_sf, palm_buffer)) > 0)

# Verify the new binary overlap
print(table(palm_model_data$weevil_overlap))

# Convert to factor
palm_model_data$weevil_overlap <- factor(palm_model_data$weevil_overlap)

```

Then fit a Generalised Linear Model (GLM) to assess whether the presence of Red Palm Weevil within a 10 km buffer affects the distribution of Date Palm, while also accounting for climatic variables (BIO5, BIO6, BIO14, BIO16).

```{r}
# Fit the GLM with weevil overlap as a predictor
glm_biotic_dependence <- glm(
  presence ~ bio5 + bio6 + bio14 + bio16 + weevil_overlap,
  data = palm_model_data,
  family = binomial(link = "logit")
)

# View model summary
summary(glm_biotic_dependence)

```
_..._

The Date Palm distribution model shows that temperature and precipitation are key factors influencing its presence. BIO5 (Max Temperature of the Warmest Month) has a negative effect on palm occurrence, indicating sensitivity to heat. BIO6 (Min Temperature of the Coldest Month) positively influences palm survival, reflecting its tolerance to milder winters. BIO14 (Precipitation of the Driest Month) negatively impacts palm presence, highlighting the species' need for water availability. BIO16 (Precipitation of the Wettest Quarter) also shows a small negative effect.

The Weevil Overlap variable is significantly positive, suggesting that palm presence is more likely where Red Palm Weevil is found, likely due to shared environmental conditions.

The model is statistically strong, with all variables being highly significant (p < 0.001). The AIC of 1593.9 and the improvement in deviance indicate a good fit, with climatic variables playing the dominant role in shaping the Date Palm's distribution.


## Question 5

### Load Future Climate Projections
To predict how future climate changes may affect the distribution of Phoenix dactylifera (Date Palm) and Rhynchophorus ferrugineus (Red Palm Weevil), we used CMIP6 climate projections for the period 2061–2080, selecting the CanESM5 model and SSP2-4.5 scenario. This model provides future bioclimatic data that helps to estimate changes in temperature, precipitation, and other climate factors affecting the species' habitats.

The relevant bioclimatic data was either loaded from an existing file or downloaded and cropped to fit the study region.

```{r}

#| label: prepare-future-bioclim-data
#| description: "Load or download and process future CMIP6 bioclimatic raster data (2061–2080, SSP2-4.5)"

# Define file path
future_bioclim_file <- here("data", "raw", "climate", "cmip6_2061_2080_canesm5_ssp245.tif")

if (file.exists(future_bioclim_file)) {
  future_bioclim_data <- rast(future_bioclim_file)
} else {
  future_bioclim_data <- cmip6_world(
    model = "CanESM5",
    var = "bio", 
    ssp = "245",
    res = 10,
    time = "2061-2080",
    path = here("data", "raw", "climate")
  ) %>%
    magrittr::set_names(names(bioclim_data)) %>%
    crop(buffered_region)

  writeRaster(future_bioclim_data, future_bioclim_file, overwrite = TRUE)
}


```

Using the future climate projections, we predicted the future suitability of the Date Palm species in the study region. The model used for prediction is based on the Generalized Linear Model (GLM) built with current bioclimatic variables.

The output of the prediction provides the estimated occurrence probabilities for Date Palm under future climatic conditions.

```{r}

# ...

# Predict future suitability based on final model
palm_prediction_future <- predict(
  future_bioclim_data,
  final_palm_model,
  ext = study_region,
  type = "response"
) %>%
  crop(study_region)


# Source plotting functions
source(here("functions", "plotting.R"))
# 
plot_glm_probability_raster(
  final_glm_model = final_palm_model,
  climate_data = future_bioclim_data,
  buffered_region = buffered_region,
  title = "GLM Probability of Palm Occurrence (2061–2080)"
)

```
The map generated (Figure 1) illustrates the predicted distribution of Date Palm in the period 2061–2080, with areas of higher suitability marked by darker colours, indicating regions where Date Palm is expected to thrive under future climate conditions.

Step 3: Predict Future Distribution of Red Palm Weevil
Similarly, we predicted the future distribution of the Red Palm Weevil based on the future climate projections and the GLM built for the weevil. The future suitability for the Red Palm Weevil was calculated using the same procedure as for Date Palm.
Step 3: Predict Future Distribution of Red Palm Weevil
```{r}
## ---------------------------
# Step 3: Predict Future Distribution of Red Palm Weevil ----

# Predict future suitability based on final model
weevil_prediction_future <- predict(
  future_bioclim_data,
  final_weevil_model,
  ext = study_region,
  type = "response"
) %>%
  crop(study_region)

# Source plotting functions
source(here("functions", "plotting.R"))
# Plot 
plot_glm_probability_raster(
  final_glm_model = final_weevil_model,
  climate_data = future_bioclim_data,
  buffered_region = buffered_region,
  title = "GLM Probability of Weevil Occurrence (2061–2080)"
)

```
The output (Figure 2) represents the areas with the highest suitability for Red Palm Weevil under future climatic conditions. This allows us to assess potential overlap with areas predicted to be suitable for Date Palm, highlighting the potential for increased pest pressure in certain regions.

Step 4: Quantifying Range Changes
Next, we converted the predicted suitability raster data into binary values, where 1 indicates suitable habitat (i.e., where species could potentially exist) and 0 indicates unsuitable habitat. We then converted these binary rasters into point geometries to calculate changes in species' ranges.

For Date Palm and Red Palm Weevil, we calculated the areas that are predicted to remain suitable (or suitable under future conditions) by converting the binary raster data into points and creating buffers around these points.
```{r}
#| label: prepare-future-buffered-points
#| description: "Threshold predictions, convert to sf points, project and buffer palm range"

# Define threshold
threshold <- 0.5

# Convert raster predictions to binary (1 = suitable)
palm_binary_future   <- palm_prediction_future >= threshold
weevil_binary_future <- weevil_prediction_future >= threshold

# Convert binary rasters to point geometries
palm_future_points <- as.points(palm_binary_future, values = TRUE, na.rm = TRUE)

# Get the actual layer name (first layer in raster)
layer_name <- names(palm_binary_future)[1]

# Subset where value == 1 (suitable habitat)
palm_future_points <- palm_future_points[palm_future_points[[layer_name]] == 1, ]

# Convert to sf object
palm_future_points <- st_as_sf(palm_future_points, crs = 4326)


weevil_future_points <- as.points(weevil_binary_future, values = TRUE, na.rm = TRUE)
layer_name_weevil <- names(weevil_binary_future)[1]
weevil_future_points <- weevil_future_points[weevil_future_points[[layer_name_weevil]] == 1, ]
weevil_future_points <- st_as_sf(weevil_future_points, crs = 4326)

# Reproject to metric CRS for buffering
crs_proj <- 3857
palm_future_proj   <- st_transform(palm_future_points, crs = crs_proj)
weevil_future_proj <- st_transform(weevil_future_points, crs = crs_proj)

# Create 10 km buffer around future palm points
palm_future_buffer <- st_buffer(palm_future_proj, dist = 10000)

```


Step 5: Percentage Change in Species Range
Finally, we calculated the percentage change in the spatial overlap between Date Palm and Red Palm Weevil from the present to future conditions. The percentage change was determined by comparing the number of weevil presence points within a 10 km buffer around the future suitable Date Palm habitat with the current overlap.

```{r}
#| label: calculate-overlap-percent-change
#| description: "Calculate percentage change in palm–weevil overlap from present to future"

# Calculate current number of weevils inside the palm buffer
weevils_in_buffer <- weevil_proj %>% 
  st_within(palm_buffer, sparse = FALSE) %>%  # Check which weevils fall within the buffer
  rowSums() > 0  # Logical vector: TRUE if within the buffer

# Calculate percentage of future weevils inside future palm buffer
future_in_buffer <- weevil_future_proj %>%
  st_within(palm_future_buffer, sparse = FALSE) %>%  # Spatial join: which weevils fall within the future buffer
  rowSums() > 0  # Logical vector: TRUE if within any buffer

# Calculate percent change compared to present
overlap_change_percent <- ((sum(future_in_buffer) - sum(weevils_in_buffer)) / sum(weevils_in_buffer)) * 100

# Print the percentage change
print(overlap_change_percent)

```
The results showed that the proportion of weevil presence points falling within the palm buffer zone is expected to increase in the future. This suggests that Red Palm Weevil is likely to expand its range into regions with suitable Date Palm habitats under future climate conditions.

## Conclusion
The projected distribution of Date Palm under future climate conditions indicates a slight contraction in suitable habitat, whereas the Red Palm Weevil is expected to either maintain or expand its range. As a result, the overlap between the two species is expected to increase, with a higher proportion of weevil presences occurring within 10 km of suitable Date Palm habitat.

These findings suggest that, under future climate scenarios, host-pest interactions between Phoenix dactylifera and Rhynchophorus ferrugineus could intensify. Increased co-occurrence may exacerbate the risk of pest outbreaks in regions suitable for Date Palm cultivation, emphasizing the need for proactive pest management strategies as climate conditions change.

